Model Search in Relational Domains

### 2016-10-23 background.tex

\paragraph{Related Work}
The work of~\cite{sas13,DBLP:journals/fmsd/BrainDGHK14} present an
instantiation of ACDCL as a decision procedure for floating-point
arithmetic that uses interval constraint propagation.  
%
The use of
ACDCL as program analysis is proposed in~\cite{tacas12}.  The analysis
operates over the abstract domain of floating-point numbers and
machine integers.  Thus, it is restricted to bounds checking for
numerical programs.  
%
A similar technique that lifts DPLL(T) to
programs is Satisfiability Modulo Path Programs (SMPP) \cite{SMPP}. SMPP
enumerates program paths using a SAT formula, which are then verified
using abstract interpretation.  
%
The lifting of CDCL to first-order theories is proposed
in~\cite{dpll,cp09,ndsmt}.
%
\Omit{ operates on a fixed first-order partial assignment lattice
  structure, where first-order variables are mapped to domain values,
  similar to constants lattice in program analysis.  } However, unlike
previous works that operate on a fixed first-order lattice structure,
ACDCL can be instantiated with different abstract domains.  This
involves model search and learning in abstract lattices.  \Omit { The
  abstract lattice in natural domain SMT does not have complementable
  meet irreducibles, and therefore does not support generalized clause
  learning~\cite{sas}.  On the other hand, most abstract lattice have
  complementation property, thus enabling ACDCL to perform generalized
  clause learning.  } 
%
\cite{DBLP:journals/fmsd/BrainDGHK14} presents an instantiation of ACDCL for
heap-manipulating programs, but uses a simplistic conflict analysis.
%
\cite{DBLP:conf/vmcai/PelleauMTB13} describes a constraint solving algorithm
using relational numerical domains, but performs model search only.
%
\cite{DBLP:conf/esop/MineBR16} proposes an algorithm inspired by constraint
solvers for inferring disjunctive invariants using the interval domain.
%
% is presented in . They
%use a relational domain, but focus on the model search, whereas the
%conflict analysis is simplistic and restarts the algorithm with the
%learnt clause instead of backtracking.
%}

>>>>

\section{Related Work}
The work of~\cite{sas13,DBLP:journals/fmsd/BrainDGHK14} present an
instantiation of ACDCL as a decision procedure for floating-point
arithmetic that uses interval constraint propagation.  
%
The use of
ACDCL as program analysis is proposed in~\cite{tacas12}.  The analysis
operates over the abstract domain of floating-point numbers and
machine integers.  Thus, it is restricted to bounds checking for
numerical programs.  
%
A similar technique that lifts DPLL(T) to
programs is Satisfiability Modulo Path Programs (SMPP) \cite{SMPP}. SMPP
enumerates program paths using a SAT formula, which are then verified
using abstract interpretation.  
%
The lifting of CDCL to first-order theories is proposed
in~\cite{dpll,cp09,ndsmt}.
%
\Omit{ operates on a fixed first-order partial assignment lattice
  structure, where first-order variables are mapped to domain values,
  similar to constants lattice in program analysis.  } However, unlike
previous works that operate on a fixed first-order lattice structure,
ACDCL can be instantiated with different abstract domains.  This
involves model search and learning in abstract lattices.  \Omit { The
  abstract lattice in natural domain SMT does not have complementable
  meet irreducibles, and therefore does not support generalized clause
  learning~\cite{sas}.  On the other hand, most abstract lattice have
  complementation property, thus enabling ACDCL to perform generalized
  clause learning.  } 
%
\cite{DBLP:journals/fmsd/BrainDGHK14} presents an instantiation of ACDCL for
heap-manipulating programs, but uses a simplistic conflict analysis.
%
\cite{DBLP:conf/vmcai/PelleauMTB13} describes a constraint solving algorithm
using relational numerical domains, but performs model search only.
%
\cite{DBLP:conf/esop/MineBR16} proposes an algorithm inspired by constraint
solvers for inferring disjunctive invariants using the intervals.
%
% is presented in . They
%use a relational domain, but focus on the model search, whereas the
%conflict analysis is simplistic and restarts the algorithm with the
%learnt clause instead of backtracking.
%}

### 2016-10-23 example.tex

BMC converts the program into a bit-vector equation and passes that to a
CDCL-based SAT solver.  We will contast the analysis performed by CDCL with
that of ACDCL.  The right-hand side of Figure~\ref{fig:example} illustrates
an analysis using ACDCL with the octagon domain (top) and the interval
domain (bottom). The analysis associates an abstract element with each
control location and variable.  The octagons deduced by ACDCL are sufficient
to prove safety without any decision or clause learning, as in the case of
Ast{\'e}e.

The benefit of ACDCL on this example becomes apparent using the interval
domain.  The intervals generated by forward analysis in the initial
deduction phase are $\langle x:[-5,5], z:[-25,25] \rangle$.  Clearly, these
do not prove safety.  Hence, ACDCL makes a heuristic decision to refine the
analysis.  Assume we decide that $c$ is $[1,1]$.  Using deduction, ACDCL
determines that the interval of $x$ is then constrained to $[0,5]$.  The
deductions made during fixed point iteration are represented by an abstract
conflict graph, given in right-hand side of Figure~\ref{fig:example}.  Nodes
in the abstract conflict graph denote updates to the program variables in
the CFG.  The interval analysis concludes that $(\text{Error}:\bot)$, that
is, the decision $c:=[1,1]$ leads to a {\em conflict}.  Thus, the program is
{\em safe} for the case $c:=[1,1]$.

At this point, a clause learning propositional SAT solver learns a reason
for the conflict and then backtracks to a level such that the learnt clause
is \emph{unit}.  Similarly, ACDCL learns that $c:=[0,0]$ at node $n0$, that
is, all error traces must satisfy $c \neq 1$ at node~$n0$.  The analysis
discards all interval constraints that lead to the conflict and backtracks
to decision level~0.  ACDCL then performs interval analysis with the learnt
constraint $(c \neq 1)$, which also leads to a conflict, as shown in
right-hand side of Figure~\ref{ssa}.  The analysis cannot backtrack further
and therefore it terminates, proving that the program is safe.  Thus,
decision and clause learning are used to avoid case-based reasoning and
prevent enumeration behaviour.

>>>>

BMC converts the program into a bit-vector equation and passes that to a
CDCL-based SAT solver.  We will contast the analysis performed by CDCL with
that of ACDCL.  The right-hand side of Figure~\ref{fig:example} illustrates
an analysis using ACDCL with the octagon domain (top) and the interval
domain (bottom). The analysis associates an abstract element with each
control location.  The octagons deduced by ACDCL are sufficient
to prove safety without any decision or clause learning, as in the case of
Astr{\'e}e.

The benefit of ACDCL on this example becomes apparent using the interval
domain.  The intervals generated by forward analysis in the initial
deduction phase are $\langle x:[-5,5], z:[-25,25] \rangle$.  Clearly, these
do not prove safety.  Hence, ACDCL makes a heuristic decision to refine the
analysis.  Assume we decide that $c$ is $[1,1]$.  Using deduction, ACDCL
determines that the interval of $x$ is then constrained to $[0,5]$.  The
deductions made during fixed point iteration are represented by an abstract
conflict graph, given in right-hand side of Figure~\ref{fig:example}.  Nodes
in the abstract conflict graph denote updates to the program variables in
the CFG.  The interval analysis concludes that $(\text{Error}:\bot)$, that
is, the decision $c:=[1,1]$ leads to a {\em conflict}.  Thus, the program is
{\em safe} for the case $c:=[1,1]$.
%
At this point, a clause learning propositional SAT solver learns a reason
for the conflict and then backtracks to a level such that the learnt clause
is \emph{unit}.  Similarly, ACDCL learns that $c:=[0,0]$ at node $n0$, that
is, all error traces must satisfy $c \neq 1$ at node~$n0$.  The analysis
discards all interval constraints that lead to the conflict and backtracks
to decision level~0.  ACDCL then performs interval analysis with the learnt
constraint $(c \neq 1)$, which also leads to a conflict, as shown in
right-hand side of Figure~\ref{ssa}.  The analysis cannot backtrack further
and therefore it terminates, proving that the program is safe.  Thus,
decision and clause learning are used to avoid case-based reasoning and
prevent enumeration behaviour.

### 2016-10-23 introduction.tex

%%%%%%%%%%%%%%%%%%%%%%% AI %%%%%%%%%%%%%%%%%%%%%%%%
%
Static program analysis based on abstract interpretation~\cite{CC77} has
been widely used to verify of properties of safety-critical systems. 
Standard static analyses~\cite{se2011} aim to compute program invariants as
fixed-points of abstract transformers.  Abstract states are chosen from a
lattice that has meet and join operations.  The meet precisely models set
intersection (or conjunction, taking a logical view), and the join
over-approximates set union (or disjunction).  The over-approximation in the
join operation is a key cause of loss of precision, which yields false
alarms.  Standard means to address false alarms in static analysis are
richer abstract domains or equipping the analysis with more precise set
unions -- which is often very expensive since the analysis exhibits case
enumeration behaviour.

>>>>

%%%%%%%%%%%%%%%%%%%%%%% AI %%%%%%%%%%%%%%%%%%%%%%%%
%
Static program analysis based on abstract interpretation~\cite{CC77} has
been widely used to verify of properties of safety-critical systems. 
Standard static analyses~\cite{se2011} aim to compute program invariants as
fixed-points of abstract transformers.  Abstract states are chosen from a
lattice that has meet and join operations.  The meet precisely models set
intersection (or conjunction, taking a logical view), and the join
over-approximates set union (or disjunction).  The over-approximation in the
join operation is a key cause of loss of precision, which yields false
alarms.  Standard means to address false alarms in static analysis are
richer abstract domains or equipping the analysis with more precise set
unions -- which is often expensive since the analysis exhibits case
enumeration behaviour.

### 2016-10-23 learning.tex
formatting of aunit

\section{Abstract Conflict Analysis}
Conflict analysis in a SAT solver can be seen as abductive 
reasoning which underapproximates a set of models that do not satisfy a 
formula~\cite{sas12,dhk2013-popl}.  
\Omit {
where a singleton assignment $c$ is replaced by a partial 
assignment that is sufficient to infer $c$.  Thus, a conflict 
analysis is a generalization of an under-approximation of a set of 
countermodels, where a countermodel is a 
set of structures that do not satisfy a formula 
}
Since relational abstract domains have more complex structure than the partial 
assignments domain or interval domain, so finding a generalized reason for 
conflict through UIP computation~\cite{uip} is non-trivial.  We formally 
describe the adaptations of conflict analysis procedure for relational domains. %
\begin{figure}[t]
%\caption{An example of Octagon}\label{octagon}
\scalebox{.65}{\import{figures/}{conflict_graph.pspdftex}}
\caption{\label{conflict} Abstract Conflict Graph}
\end{figure}

@@

\paragraph {\textbf{Lifting UIP to relational domains}}
An UIP is a special node in the abstract conflict graph such that 
any path from the last decision node to the conflict node must pass 
through it.  Unlike a SAT solver, a UIP for relational domain may 
involve meet irreducibles that contain the same variables.  The 
first-UIP is a unique node closest to the conflict, and the last UIP 
is the decision node itself.  Computing UIPs ensures asserting
cuts~\cite{cdcl,DBLP:journals/fmsd/BrainDGHK14}, that is, it 
yields clauses that generate new deductions after backtracking.  
Every cut in the graph is a reason for conflict that can be 
used in learning.  An abstract first-UIP algorithm~\cite{DBLP:journals/fmsd/BrainDGHK14} 
traverses the trail $\trail$ starting from the conflict node and 
computes a cut that suffices to produce a conflict. 
\Omit {
contradict with the transformer stored in $\reasons[\bot]$.  
}
For our example, there exist multiple incomaparable reasons for conflict,
marked as {\em cut0, cut1}, in Figure~\ref{conflict}.  Here, cut0 is the first UIP.  
Choosing cut0 yields a learnt clause 
$(x+y<15 \vee -x-y<-15 \vee -x+y<5 \vee x-y<-5 \vee y-10<0 \vee -y+10<0)$, 
which is obtained by negating the reason for conflict.  
%    
\paragraph{\textbf{Learning in relational domains}}
Learning in a propositional solver yields an asserting
clause~\cite{cdcl} that expresses the negation of the conflict
reasons.  However, we model learning in relational domains as learning
a transformer which infers new meet irreducibles using the abstract
unit rule called {\em abstract unit transformer}.  We add $\aunit$ to
the set of abstract transformers $\abstransset$ used during model
search. $\aunit$ is a generalization of propositional unit rule for
numerical domains.  For an abstract lattice $\domain$ with
complementable meet irreducibles and a set of meet irreducibles $\conflictset
\subseteq \domain$ such that $\bigsqcap
\conflictset$ does not satisfy $\formula$, $\aunit_\conflictset: \domain \rightarrow
\domain$ is formally defined as follows.
\[ \aunit(\absval) =
 \begin{cases}
  \bot       & \quad \text{if } \absval \sqsubseteq \bigsqcap \conflictset \qquad \qquad \qquad
  \qquad \qquad \qquad \quad (1)\\
  \bar{t}    & \quad \text{if } t \in \conflictset \; \text{and} \; \forall t' \in \conflictset
  \setminus \{t\}. \absval  \sqsubseteq t' \qquad \quad \; \; (2) \\
  \bar{t}    & \quad \text{if } t \in \conflictset \; \text{and} \; \forall t' \in \conflictset \setminus \{t\}. \absval
  \not\sqsubseteq t', \; \text{then} \\ 
             & \quad \forall \absval' \sqsubseteq \absval: \absval'
             \not\sqsubseteq t \text{ and}\\ 
             & \quad \absval''=\exclude(\absval,\absval') \text{ and
             }\absval''\sqsubseteq t' \qquad  \; \; (3) \\
  \top & \quad \text{otherwise} \qquad \qquad \qquad \qquad \qquad \qquad \quad (4) \\
 \end{cases}
\]
with $\exclude(\absval,\absval')=\bigsqcap(\decomp(\absval) \setminus
\decomp(\absval'\}))$.  $\aunit$ returns $\bot$ if $\conflictset$ is
conflicting following rule (1).  Rule (2) and (3) of $\aunit$ infer a
valid meet irreducible, which implies that $\conflictset$ is unit.

@@

Let us consider another example, where $\conflictset = \{x \geq 2, x
\leq 5, y \leq 7 \}$ and let $\absval = (x \geq 3\wedge x \leq 4\wedge
y \geq 5\wedge y \leq 6)$, then $\aunit_\conflictset(\absval) = \bot$
using the rule (1), since $\absval \sqsubseteq \bigsqcap\conflictset$.  Now,
let $\absval' = (x \geq 3\wedge x \leq 4)$, then
$\aunit_\conflictset(\absval') = (y \geq 8)$ using rule (2), since
$\absval' \sqsubseteq (x \geq 2)$ and $\absval' \sqsubseteq (x \leq
5)$.  However, if none of the above rule holds true, $\aunit$ returns~$\top$.

>>>>

\section{Abstract Conflict Analysis}\label{sec:conflict}
Conflict analysis in a SAT solver can be seen as abductive 
reasoning which underapproximates a set of models that do not satisfy a 
formula~\cite{sas12,dhk2013-popl}.  
\Omit {
where a singleton assignment $c$ is replaced by a partial 
assignment that is sufficient to infer $c$.  Thus, a conflict 
analysis is a generalization of an under-approximation of a set of 
countermodels, where a countermodel is a 
set of structures that do not satisfy a formula 
}
Since relational abstract domains have more complex structure than the partial 
assignments domain or interval domain, so finding a generalized reason for 
conflict through UIP computation~\cite{uip} is non-trivial.  We formally 
describe the adaptations of conflict analysis procedure for relational domains. %
\begin{figure}[t]
\vspace*{-10ex} %BAD TRICK
%\caption{An example of Octagon}\label{octagon}
\scalebox{.65}{\import{figures/}{conflict_graph.pspdftex}}
\caption{\label{conflict} Abstract Conflict Graph}
\end{figure}

@@

\paragraph {\textbf{Lifting UIP to relational domains}}
An UIP is a node in the abstract conflict graph such that 
any path from the last decision node to the conflict node must pass 
through it.  Unlike a SAT solver, a UIP for a relational domain may 
involve meet irreducibles that contain the same variables.  The 
first UIP is a unique node closest to the conflict, and the last UIP 
is the decision node itself.  Computing UIPs ensures asserting
cuts~\cite{cdcl,DBLP:journals/fmsd/BrainDGHK14}, that is, it 
yields clauses that generate new deductions after backtracking.  
Every cut in the graph is a reason for conflict that can be 
used in learning.  An abstract first-UIP algorithm~\cite{DBLP:journals/fmsd/BrainDGHK14} 
traverses the trail $\trail$ starting from the conflict node and 
computes a cut that suffices to produce a conflict. 
\Omit {
contradict with the transformer stored in $\reasons[\bot]$.  
}
For our example, there exist multiple incomaparable reasons for conflict,
marked as {\em cut0, cut1}, in Figure~\ref{conflict}.  Here, cut0 is the first UIP.  
Choosing cut0 yields a learnt clause 
$(x+y<15 \vee -x-y<-15 \vee -x+y<5 \vee x-y<-5 \vee y-10<0 \vee -y+10<0)$, 
which is obtained by negating the reason for conflict.  
%    
\paragraph{\textbf{Learning in relational domains}}
Learning in a propositional solver yields an asserting
clause~\cite{cdcl} that expresses the negation of the conflict
reasons.  However, we model learning in relational domains as learning
a transformer which infers new meet irreducibles using the abstract
unit rule called {\em abstract unit transformer}.  We add $\aunit$ to
the set of abstract transformers $\abstransset$ used during model
search. $\aunit$ is a generalization of propositional unit rule for
numerical domains.  For an abstract lattice $\domain$ with
complementable meet irreducibles and a set of meet irreducibles $\conflictset
\subseteq \domain$ such that $\bigsqcap
\conflictset$ does not satisfy $\formula$, $\aunit_\conflictset: \domain \rightarrow
\domain$ is formally defined as follows.
\[ \aunit(\absval) =
 \left\{\begin{array}{l@{\quad}l@{\qquad}l}
  \bot       & \text{if } \absval \sqsubseteq \bigsqcap \conflictset & (1)\\
  \bar{t}    & \text{if } t \in \conflictset \; \text{and} \; \forall t' \in \conflictset
  \setminus \{t\}. \absval  \sqsubseteq t' & (2) \\
  \bar{t}    & \text{if } t \in \conflictset \; \text{and} \; \forall t' \in \conflictset \setminus \{t\}. \absval
  \not\sqsubseteq t', \; \text{then} \\ 
             & \quad \forall \absval' \sqsubseteq \absval: \absval'
             \not\sqsubseteq t \text{ and} \\ 
             & \quad \absval''=\exclude(\absval,\absval') \text{ and
             }\absval''\sqsubseteq t' & (3) \\
  \top & \text{otherwise} & (4) \\
 \end{array}\right.
\]
with $\exclude(\absval,\absval')=\bigsqcap(\decomp(\absval) \setminus
\decomp(\absval'\}))$.  $\aunit$ returns $\bot$ if $\conflictset$ is
conflicting following rule (1).  Rule (2) and (3) of $\aunit$ infer a
valid meet irreducible, which implies that $\conflictset$ is unit.

@@

Let us consider another example, where $\conflictset = \{x \geq 2, x
\leq 5, y \leq 7 \}$ and let $\absval = (x \geq 3\wedge x \leq 4\wedge
y \geq 5\wedge y \leq 6)$, then $\aunit_\conflictset(\absval) = \bot$
using the rule (1), since $\absval \sqsubseteq \bigsqcap\conflictset$.  Now,
let $\absval' = (x \geq 3\wedge x \leq 4)$, then
$\aunit_\conflictset(\absval') = (y \geq 8)$ using rule (2), since
$\absval' \sqsubseteq (x \geq 2)$ and $\absval' \sqsubseteq (x \leq
5)$.  However, if none of the above rules hold true, $\aunit$ returns~$\top$.
